---
title: "Lab 19: Flu Finder"
author: "Abigail Perez & German Vega"
format: pdf
editor: visual
---

```{r}
#| eval: false
install.packages("seqinr")
install.packages("stringr")
```

```{r}
upload_fasta <- function(fasta_filename) {
  library(seqinr)
  read.fasta(fasta_filename, seqtype = "AA", as.string = TRUE,
             set.attributes = FALSE)
}
```

```{r}
upload_fasta("testfasta.txt")       # should return list of proteins A, B, C
upload_fasta("testfasta.txt")$A     # should return just protein A's sequence
```

```{r}
trypsinize <- function(proteins) {
  library(stringr)
  lapply(proteins, str_split_1, pattern="(?<=R|K)")
}
```

```{r}
proteins <- list(A="LVKLHHIIFESMLKDMQRRHRVW", 
                 B="ADEFQGSMQKIEACWQSYDVQF", 
                 C="MINEPFSWRLEFHLSERKYDEIM")
trypsinize(proteins)
trypsinize(proteins)$A
```

```{r}
split_peptides <- function(peptides) {
  library(stringr)
  lapply(peptides, str_split, pattern="")
}
```

```{r}
peptides <- list(A=c("LVK","LHHIIFESMLK","DMQR","R","HR","VW"), 
                 B=c("ADEFQGSMQK","IEACWQSYDVQF"), 
                 C=c("MINEPFSWR","LEFHLSER","K","YDEIM"))
split_peptides(peptides)
split_peptides(peptides)$B
```

```{r}
splitpeptides_to_masses <- function(aa) {
  aa_masses <- c(A=71.037, R=156.101, N=114.042, D=115.026, C=103.009, 
                 Q=128.058, E=129.042, G=57.021, H=137.058, I=113.084, 
                 L=113.084, K=128.094, M=131.040, F=147.068, P=97.052, 
                 S=87.032, T=101.047, W=186.079, Y=163.063, V=99.068)
  
  peptide_masses <- aa
  for(i in 1:length(aa)) {
    peptide_masses[[i]] <- lapply(aa[[i]], function(x) sum(aa_masses[x]))
  }
  lapply(peptide_masses, unlist)
}
```

```{r}
aa <- list(A=list(c("L","V","K"), 
                  c("L","H","H","I","I","F","E","S","M","L","K"), 
                  c("D","M","Q","R")), 
           B=list(c("A","D","E","F","Q","G","S","M","Q","K"), 
                  c("I","E","A","C","W","Q","S","Y","D","V","Q","F")), 
           C=list(c("M","I","N","E","P","F","S","W","R"), 
                  c("L","E","F","H","L","S","E","R"), 
                  c("Y","D","E","I","M")))
splitpeptides_to_masses(aa)
splitpeptides_to_masses(aa)$A    # expect: 340.246  1348.728  530.225
```

```{r}
count_matching_masses <- function(protein_masses, sample) {
  df <- as.data.frame(sapply(protein_masses, function(x) 
    sum(as.character(sample) %in% as.character(x))))
  names(df) <- "peptide_counts"
  return(df)
}
```

```{r}
masses_list <- list(A=c(340.246, 1348.728, 530.225), 
                    B=c(1121.476, 1469.624), 
                    C=c(1160.540, 1011.511, 651.255))
sample <- c(340.246, 530.225, 1348.728)
count_matching_masses(masses_list, sample)
# expect: A=3, B=0, C=0

# Then test with a new random sample:
sample2 <- c(1121.476, 1160.540, 340.246)
count_matching_masses(masses_list, sample2)
```

```{r}
ggbarplot <- function(peptide_counts_table) {
  library(ggplot2)
  ggplot(peptide_counts_table) +
    aes(rownames(peptide_counts_table), peptide_counts) +
    geom_col(fill="maroon2", width=0.5) +
    theme_bw() +
    labs(x="Flu Strain", y="Peptide Counts")
}
```

```{r}
counts_df <- data.frame(peptide_counts=c(3, 0, 0), rownames=c("A", "B", "C"))
ggbarplot(counts_df)

# Also test with different numbers:
counts_df2 <- data.frame(peptide_counts=c(1, 2, 5), rownames=c("A", "B", "C"))
ggbarplot(counts_df2)
```
```{r}
list.files(".")          # verify flufinder.R is there
source("flufinder.R")    # loads all 6 functions
```

```{r}
# Load patient sample
patient_sample <- scan("patient_sample.txt")

# Step 1-2: upload and trypsinize
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize()

# Step 1-4: add splitting and mass calculation
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |> 
  split_peptides() |> splitpeptides_to_masses()

# Step 1-5: count matches against patient sample
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |> 
  split_peptides() |> splitpeptides_to_masses() |> 
  count_matching_masses(sample=patient_sample)

# Full pipeline with plot
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |> 
  split_peptides() |> splitpeptides_to_masses() |> 
  count_matching_masses(sample=patient_sample) |> ggbarplot()
```

```{r}
list.files("/Users/abbyperez/Library/Mobile Documents/com~apple~CloudDocs/BIMM 143/Flu Project")

source("/Users/abbyperez/Library/Mobile Documents/com~apple~CloudDocs/BIMM 143/Flu Project/flufinder.R")
```


```{r}
# Using scan() to generate a vector from a text file with space separated numbers
patient_sample <- scan("patient_sample.txt")
patient_sample
```


```{r}
# Inspecting the flu protein database
upload_fasta("flu_proteins_database.txt")
```


```{r}
# Piping the flu protein database to trypsinize()
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize()
```


> **Q1: What do the output strings represent and why do they all end in R or K?**

The output strings are tryptic peptide fragments from each flu strain's HA protein. They all end in R or K because trypsin cleaves specifically after those two amino acids, so every fragment (except the very last one) ends with either K or R.

```{r}
# Piping until splitpeptides_to_masses()
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |>
split_peptides() |> splitpeptides_to_masses()
```


> **Q2: What do all the numbers represent?**

The numbers are the calculated molecular masses (in Daltons) of each tryptic peptide, derived by summing the individual amino acid masses for each fragment.

```{r}
# Piping until count_matching_masses()
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |>
split_peptides() |> splitpeptides_to_masses() |>
count_matching_masses(sample=patient_sample)
```


> **Q3: What do the numbers in the table represent and why does the count_matching_masses() function need the sample input in addition to the standard input from the pipeline?**

The numbers represent how many peptide masses from each flu strain match the masses detected in the patient's mass spec sample. count_matching_masses() needs the patient sample as a separate input because it's not part of the database pipeline — it's the external reference we're comparing against to identify the strain.

```{r}
# Using piping to consecutively run all functions of the flufinder script
"flu_proteins_database.txt" |> upload_fasta() |> trypsinize() |>
split_peptides() |> splitpeptides_to_masses() |>
count_matching_masses(sample=patient_sample) |> ggbarplot()
```


> **Q4: Which flu strain do you order an mRNA vaccine to be mass produced for?**

H1N1 — it has by far the most matches (35), so that's the strain to order the vaccine for.

> **Q5: Why do the other flu strains show a subset of matches?**

All flu strains share some evolutionarily conserved regions in their HA proteins, so a few peptides happen to have identical masses across strains. Those shared peptides show up as low-count matches (3–6) in the other strains, but only H1N1 matches nearly the entire patient sample.
